% =============================================================================
% AggreQuant: Methods Paper
% Main Document
% =============================================================================

\documentclass[11pt,a4paper]{article}

% --- Packages ---
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage[margin=2.5cm]{geometry}
\usepackage{natbib}
\usepackage{xcolor}
\usepackage{listings}

% --- Code listing style ---
\lstset{
    language=Python,
    basicstyle=\ttfamily\small,
    keywordstyle=\color{blue},
    commentstyle=\color{gray},
    stringstyle=\color{red},
    frame=single,
    breaklines=true
}

% --- Hyperref setup ---
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue
}

% --- Title ---
\title{AggreQuant: A Python Toolkit for Quantitative Analysis of \\
       Fluorescence Microscopy Images}

\author{
    Athena Economides\textsuperscript{1,*} \\[1em]
    \textsuperscript{1}University of Zurich, Switzerland \\
    \textsuperscript{*}Corresponding author
}

\date{\today}

% =============================================================================
\begin{document}
% =============================================================================

\maketitle

% -----------------------------------------------------------------------------
\begin{abstract}
AggreQuant is an open-source Python toolkit for automated analysis of fluorescence microscopy images, with particular focus on quality control and quantitative feature extraction. The toolkit implements state-of-the-art methods for image focus quality assessment, combining spatial domain metrics (Variance of Laplacian, Brenner gradient, Sobel-based measures) with frequency domain analysis (Power Log-Log Slope) to enable robust detection of both localized blur regions and global defocus. We provide comprehensive benchmarking against established methods and demonstrate that the combination of PLLS for global assessment and patch-based VoL for local blur detection provides effective quality control without requiring deep learning models or training data.
\end{abstract}

% -----------------------------------------------------------------------------
\section{Introduction}
\label{sec:introduction}

% TODO: Write introduction covering:
% - Importance of automated microscopy image analysis
% - Challenges in high-throughput imaging (quality control, variability)
% - Overview of AggreQuant capabilities
% - Paper organization

High-throughput fluorescence microscopy has become an essential tool in biological research, enabling large-scale studies of cellular phenotypes, drug responses, and disease mechanisms. However, the acquisition of thousands to millions of images introduces significant challenges in data quality and analysis standardization. Common issues include out-of-focus images due to autofocus failures, uneven illumination, optical aberrations, and sample-related artifacts.

Robust quality control is critical for ensuring that downstream analyses---including segmentation, feature extraction, and statistical inference---are based on reliable data. Manual inspection of large image datasets is impractical, necessitating automated quality assessment methods.

AggreQuant addresses these challenges by providing a comprehensive Python toolkit for fluorescence microscopy image analysis, with particular emphasis on quality control. This paper describes the image focus quality assessment module, presenting a literature review of established methods, implementation details, and benchmark results demonstrating the effectiveness of our approach.

% -----------------------------------------------------------------------------
% Include the focus quality section
\input{focus_quality_section}

% -----------------------------------------------------------------------------
\section{Conclusion}
\label{sec:conclusion}

% TODO: Write conclusion

We have presented the focus quality assessment capabilities of AggreQuant, demonstrating that a combination of frequency domain (PLLS) and spatial domain (VoL) metrics provides robust detection of both global defocus and localized blur regions. Our benchmark results show that PLLS correctly identifies globally defocused images that spatial metrics alone fail to detect, while patch-based VoL remains valuable for detecting localized aberrations.

The implemented methods achieve performance comparable to deep learning approaches for most practical applications, without requiring training data or specialized computational resources. This makes AggreQuant's quality control module accessible to researchers across diverse computational environments.

Future work will focus on expanding the quality control capabilities to include detection of other artifacts such as debris, saturation, and illumination non-uniformity, as well as integration with automated image rejection pipelines.

% -----------------------------------------------------------------------------
\section*{Acknowledgments}

% TODO: Add acknowledgments

% -----------------------------------------------------------------------------
\section*{Data and Code Availability}

AggreQuant is available as open-source software at \url{https://github.com/aecon/AggreQuant}. Benchmark images and analysis scripts are included in the repository.

% -----------------------------------------------------------------------------
\bibliographystyle{unsrtnat}
\bibliography{references}

% =============================================================================
\end{document}
% =============================================================================
